---
title: "Inferential Data Analysis"
author: "Anthony Perez Eisenbarth"
geometry: margin=1cm
output:
  pdf_document:
  template: template.tex
keep_tex: yes
latex_engine: pdflatex
header-includes:
  - \usepackage{hyperref}
  - \usepackage{array}   
  - \usepackage{caption}
  - \usepackage[flushleft]{threeparttable}
  - \usepackage[sfdefault, condensed]{roboto}

---

# Basic inferential data analysis
The goal of the following analysis of the 'ToothGrowth' data by Galton, is to quantifiably compare the effects of Vitamin C  in tooth growth of Guinea Pigs between the groups defined by two factors. The 'ToothGrowth' data consists of 60 independent observations for the length of tooth in Guinea Pigs 
over two factors, the supply method and the dose of Vitamin C. The supply method refers to how Vitamin C was supplied either as *Orange Juice* (OJ) or as *Ascorbic Acid* (VC). 


```{r, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = FALSE, message = FALSE)

library(knitr)
library(ggplot2)
library(tinytex)
library(ggthemes)
library(boot)
library(kableExtra)
library(ggthemr)
library(tidyverse)
library(dplyr)
library(scales)
library(tidyr)
library(patchwork)
```

```{r theme, include = FALSE}
raspberry <- "#DB2955"
candy_green <- "#95D7AE"
onyx <- "#313435"
babyblue <- "#47AFFF"
prussian <- "#113255"
sapphire <- "#255F85"
turq <- "#76E7CD"
emerald <- "#32936F"
violet <- "#AA78A6"
sky <- "#ABDDED"
ggthemr(palette = "fresh", layout = "clean", line_weight = 0.5)
```


```{r, include = FALSE}
data("ToothGrowth")
head(ToothGrowth)
```



```{r,  echo = FALSE, warning = FALSE}
g_len_by_supp <- ggplot(ToothGrowth, aes(supp, len)) + 
  geom_boxplot(aes(fill = supp), show.legend = FALSE) +
  theme(plot.title = element_text(face = "plain", size = 12)) +
  labs(title = "Tooth length by supplement",x = 'Supplement type', y = 'Tooth length (mm)') 
g_len_by_dose <- ggplot(ToothGrowth, aes(factor(dose), len)) + 
  geom_boxplot(aes(fill = supp)) +
  theme(plot.title = element_text(face = "plain", size = 12)) + 
  labs(x ='Dose (mg)', y = 'Tooth length (mm)', title = 'Tooth length by dose')
(g_len_by_supp | g_len_by_dose)
```


As a preliminary hypothesis, we can form the first research question around the supplement type. In general, is expected tooth length larger, when Vitamin C is supplied as orange juice instead of ascorbic acid?

```{r}
t_diff_supp <- t.test(len ~ supp, ToothGrowth, var.equal = FALSE)
t_diff_supp
```
As can be seen, considering 95% confidence the null hypothesis cannot be rejected, since the *p*-value is greater than 0.05. So we have no evidence that there is a significant difference between the two supplements.

Now, let's see if there is a significant difference between the supplements at different dosages.


```{r}
t_diff_supp_0.5 <- t.test(len ~ supp, data = subset(ToothGrowth, dose == 0.5))
t_diff_supp_0.5
```

From this result, we see that the data do provide substantial evidence 
($p < 0.006$) to reject the null hypothesis $H_0$, according to which the expected tooth length is the same when dose of 0.5 mg/day of Vitamin C is supplied as orange juice.


```{r}
t_diff_supp_1 <- t.test(len ~ supp, data = subset(ToothGrowth, dose == 1))
t_diff_supp_1
```

From this result, we see that the data do provide substantial evidence 
($p < 0.001$) to reject the null hypothesis $H_0$, according to which the expected tooth length is the same when dose of 1 mg/day of Vitamin C is supplied as orange juice.


```{r}
t_diff_supp_2 <- t.test(len ~ supp, data = subset(ToothGrowth, dose == 2))
t_diff_supp_2
```

From this result, we see that the data do *not* provide substantial evidence 
($p > 0.9638$) to reject the null hypothesis $H_0$, according to which the expected tooth length is the same when dose of 2 mg/day of Vitamin C is supplied 
either as orange juice or as ascorbic acid.


### Conclusions

The data does not provide ample evidence to support the hypothesis that there is a significant difference between Vitamin C and Asorbic acid supplements on the expected tooth length of Guinea pigs, However, when analyzing the dosage level, there is a significant difference for 0.5 and 1 mg. At both the 0.5 mg ($p < 0.006$) and 1 mg ($p < 0.001$) dosage levels, the data provides evidence to reject null hypothesis in favor the alternative that expected tooth length is larger when Vitamin C is provided as the supplement. It is not possible to say that for 2 mg dose the supplements differ in effects on expected tooth length. The data do *not* provide substantial evidence 
($p > 0.9638$) to reject the null hypothesis $H_0$, in which the expected tooth length is the same when dose of 2 mg/day of Vitamin C is supplied as orange juice.
